---
title: "Gapminder"
output: html_document
date: "2026-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Libraries
```{r}
library(tidyverse)
library(tibble)
library(janitor)
library(naniar)
library(dplyr)
library(shiny)
library(shinydashboard)
library(shinythemes)
```

#Data set and cleaning

```{r}
gapminder_life_expectancy<- read.csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/Life_expectancy.csv") 
```

```{r}
co2_emissions<-read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/co2_pcap_cons.csv")
```

```{r}
getwd()




Pop<-read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/pop.csv")
```

```{r}
Hiv_death<-read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/hiv_death.csv")
```
```{r}
Child_death<- read.csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/number_of_child_deaths.csv")

```

```{r}
#can we get rid of this from the df?

kcals_per_day<- read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/kilocalories_per_person_and_day.csv")
```

```{r}
gdp<- read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/gdp.csv")
```
```{r}
lung_cancerF<- read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/lung_cancer_f.csv")
```

```{r}
lung_cancerM<- read_csv("C:/Users/jk_ic/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/lung_cancer_m.csv")
```


#pivoting and fixing year + other problems

```{r}
gapminder_life_expectancy<- gapminder_life_expectancy %>% 
  pivot_longer(-c(name,geo),
               names_to = "year",
               values_to = "life_expectancy") %>% 
  mutate(year= str_remove(year, "X"))
```

```{r}
Child_death<-Child_death %>% 
pivot_longer(-c(name,geo),
               names_to = "year",
               values_to = "cild_death") %>% 
  mutate(year= str_remove(year, "X")) 
```

```{r}
co2_emissions<- co2_emissions %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
              values_to = "Co2_emissions")
```

```{r}
pop<- Pop %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
              values_to = "population")
```

```{r}
hiv_death<- Hiv_death %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "hiv_deaths")
```

```{r}
kcals_per_day<- kcals_per_day %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "kcal")
```

```{r}
gdp<- gdp %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "gdp")
```

```{r}
 lung_cancerF<-lung_cancerF %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "lung_cancerf")
```
  
```{r}
 lung_cancerM<-lung_cancerM %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "lung_cancerm")
```

#joining data tables

```{r}
gp<-full_join(Child_death, gapminder_life_expectancy, by= c("name", "geo","year"))
```

```{r}
gp2<-full_join(gp, co2_emissions, by= c("name", "geo","year"))
```

```{r}
gp3<-full_join(gp2,pop, by= c("name", "geo","year"))
```

```{r}
gp4<-full_join(gp3,hiv_death, by= c("name", "geo", "year") )
```

```{r}
gp5<- full_join(gp4, kcals_per_day, by = c("name","geo","year"))
```

```{r}
gp6<- full_join(gp5, gdp, by = c("name", "year","geo") )
```

```{r}
gp7<- full_join(gp6,lung_cancerF, by = c("name", "year","geo") )
```

```{r}
gapminder<- full_join(gp7,lung_cancerM, by = c("name", "year","geo") )
```


```{r}
gapminder<- gapminder %>% 
  filter(year>= "1915", year<="2022")
```

```{r}
gapminder <- gapminder %>% 
  mutate(continent = case_when(
    name %in% c("Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela") ~ "South America",
    
    name %in% c("Australia", "Papua New Guinea", "New Zealand", "Fiji", "Solomon Islands", "Vanuatu", "Samoa", "Kiribati", "Micronesia", "Tonga", "Marshall Islands", "Palau", "Nauru", "Tuvalu", "Micronesia, Fed. Sts.") ~ "Oceania",
    
    name %in% c("USA", "Mexico", "Canada", "Guatemala", "Haiti", "Dominican Republic", "Honduras", "Cuba", "Nicaragua", "El Salvador", "Costa Rica", "Panama", "Jamaica", "Trinidad and Tobago", "Belize", "Bahamas", "Barbados", "St. Lucia", "Grenada", "St. Vincent and the Grenadines", "Antigua and Barbuda", "Dominica", "St. Kitts and Nevis") ~ "North America",
    
    name %in% c("Germany", "UK", "France", "Italy","Spain", "	Ukraine", "Poland", "Romania", "Netherlands", "Belgium", "Sweden", "Czech Republic", "	Portugal", "Greece", "Hungary","Austria", "	Switzerland", "Belarus", "Bulgaria", "Serbia", "Denmark", "Norway", "Finland", "Finland", "Ireland", "Croatia", "	Bosnia and Herzegovina", "Moldova", "Lithuania", "Albania", "Slovenia", "Latvia", "North Macedonia", "Cyprus", "Estonia", "Luxembourg", "Montenegro", "Malta", "Iceland", "Andorra", "Liechtenstein", "Monaco", "San Marino", "Vatican City", "Bosnia and Herzegovina", "Switzerland", "Portugal", "Slovak Republic", "Ukraine", "Holy See") ~ "Europe",
    
    name %in% c("Russia", "Azerbaijan") ~ "Eurasia",
    
    name %in% c("India", "China", "Indonesia", "Pakistan", "Bangladesh", "Japan", "Philippines", "Vietnam", "Iran", "Turkey", "Thailand", "Myanmar", "South Korea", "Iraq", "Afghanistan", "Yemen", "Uzbekistan", "Malaysia", "Saudi Arabia", "Nepal", "North Korea", "Syria", "Sri Lanka", "Kazakhstan", "	Cambodia", "Jordan", "UAE", "Tajikistan", "	Azerbaijan", "Israel", "Laos", "Turkmenistan", "Kyrgyzstan", "Singapore", "Lebanon", "Palestine", "Oman", "Kuwait", "Georgia", "	Mongolia", "Qatar", "Armenia", "Bahrain", "Timor-Leste", "Bhutan", "	Maldives", "Brunei", "Kyrgyz Republic", "Cambodia", "Lao", "Maldives", "Mongolia") ~ "Asia", 
    
    name %in% c("Nigeria", "Ethiopia", "Egypt", "Congo, Dem. Rep.", "Tanzania", "South Africa", "Kenya", "Sudan", "Uganda", "Algeria", "Angola", "Morocco", "Mozambique", "Ghana", "Madagascar", "Cameroon", "Niger", "Mali", "Burkina Faso", "Malawi", "Zambia", "Chad", "Somalia", "Senegal", "Zimbabwe", "Guinea", "Benin", "Rwanda", "Burundi", "South Sudan", "Tunisia", "Togo", "Sierra Leone", "Libya", "Congo, Rep.", "Liberia", "Central African Republic", "Mauritania", "Eritrea", "Namibia", "Gambia", "Gabon", "Botswana", "Lesotho", "Guinea-Bissau", "Equatorial Guinea", "Eswatini", "	Mauritius", "Djibouti", "Comoros", "Cape Verde", "Sao Tome and Principe", "Seychelles", "Cote d'Ivoire", "Mauritius") ~ "Africa"
  ))
```


```{r}
gapminder <- gapminder %>% 
  mutate(longitude = case_when(
    continent %in% "Africa" ~ "19.5687",
    continent %in% "Asia" ~ "87.3311",
    continent %in% "Eurasia" ~ "70.5158",
    continent %in% "Europe" ~ "25.3200",
    continent %in% "North America" ~ "-100.1667",
    continent %in% "Oceania" ~ "133.2102",
    continent %in% "South America" ~ "-56.1015"
  ))
```

```{r}
gapminder <- gapminder %>% 
  mutate(latitude = case_when(
    continent %in% "Africa" ~ "6.614",
    continent %in% "Asia" ~ "48.35",
    continent %in% "Eurasia" ~ "49.5786",
    continent %in% "Europe" ~ "54.9067",
    continent %in% "North America" ~ "48.1667",
    continent %in% "Oceania" ~ "-25.9458",
    continent %in% "South America" ~ "-15.5991"
  ))
```

```{r}
gapminder <- gapminder %>% 
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude))
```


#final data set

```{r}
gapminder
```

```{r}
write.csv(gapminder, file="gapminder.csv", row.names=FALSE)
```

```{r}
gapminder<-read_csv("gapminder.csv")
gapminder
```

#understanding the data
```{r}
gg_miss_var(gapminder)
```

```{r}
gapminder %>% 
  miss_var_summary(order=T)
```


```{r}
gapminder %>% 
  select(name, Co2_emissions, year) %>% 
  pivot_wider(names_from = "name",
              values_from = "Co2_emissions")
```


#graphs
top 3 highest co2 emissions per capita
```{r}
gapminder %>% 
  filter(year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(avgco2=mean(Co2_emissions)) %>% 
  arrange(desc(avgco2)) %>% 
  slice_head(n=3)
  
```
bottom 3 lowest co2 emissions
```{r}
gapminder %>% 
  filter(name != c("Holy See", "Monaco", "San Marino"), year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(avgco2=mean(Co2_emissions, rm.na=T)) %>% 
  arrange(desc(avgco2)) %>% 
slice_tail(n=6)

```

6 coutnries and the lung cancer deaths

```{r}
gapminder %>% 
   filter(name ==c("Chad", "Congo, Dem. Rep.", "Burundi"), year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(mlungavg= mean(lung_cancerm), flungavg=mean(lung_cancerf)) %>% 
  ggplot(aes(x=name, y= ))
 
```

highest year of co2
```{r}
gapminder %>%
  select(name, year, Co2_emissions) %>% 
  filter(year >= "1980", year<="2022") %>% 
pivot_wider(names_from = "year",
            values_from = Co2_emissions) 
  
```

```{r}
gapminder<- gapminder %>% 
  mutate(lungcancertot=lung_cancerf+lung_cancerm)
```

co2 emissions between the 3 largest states in 2000

```{r}
gapminder %>%
  select(name, year, Co2_emissions, lung_cancerf, lung_cancerm) %>% 
  filter(name== "Qatar"|
         name == "Singapore"|
         name == "USA"|
           name== "Chad"|
           name== "Burundi"|
           name== "Congo, Dem. Rep."|
         name=="UAE",
         year== "2000") %>% 
   ggplot(aes(x=name, y=Co2_emissions, fill = name ))+
  geom_col()+
  scale_y_log10()+
  labs(title = "Co2 Emissions in the year 2000",
       x="Country",
       y= "Co2 Emissions(per capita)")
    
```



```{r}
gapminder %>%
  select(name, year, Co2_emissions, lungcancertot, lung_cancerf, lung_cancerm) %>% 
  filter(name== "Qatar"|
         name == "Singapore"|
         name == "USA"|
           name== "Chad"|
           name== "Burundi"|
         name=="UAE",
         year =="2000") %>%  #qatar, singapore, usa, chad, burundi
   pivot_longer(cols = c(lung_cancerf, lung_cancerm),
               names_to = "sex",
               values_to = "deaths")  %>% 
  ggplot(aes(x=name, y= deaths, fill = sex ))+
  geom_col(position= "dodge")+
  scale_y_log10()+
  labs(title = "Lung cancer deaths by sex in the year 2000",
       x="Country",
       y= "Deaths",
       fill= "Sex")
  

```
```{r}
gapminder
```

```{r}
gapminder %>% 
  distinct(name) %>% 
  arrange(name)
```




```{r}
gapminder <- gapminder %>% 
  mutate(continent = case_when(
    name %in% c("Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela") ~ "South America",
    
    name %in% c("Australia", "Papua New Guinea", "New Zealand", "Fiji", "Solomon Islands", "Vanuatu", "Samoa", "Kiribati", "Micronesia", "Tonga", "Marshall Islands", "Palau", "Nauru", "Tuvalu", "Micronesia, Fed. Sts.") ~ "Oceania",
    
    name %in% c("USA", "Mexico", "Canada", "Guatemala", "Haiti", "Dominican Republic", "Honduras", "Cuba", "Nicaragua", "El Salvador", "Costa Rica", "Panama", "Jamaica", "Trinidad and Tobago", "Belize", "Bahamas", "Barbados", "St. Lucia", "Grenada", "St. Vincent and the Grenadines", "Antigua and Barbuda", "Dominica", "St. Kitts and Nevis") ~ "North America",
    
    name %in% c("Germany", "UK", "France", "Italy","Spain", "	Ukraine", "Poland", "Romania", "Netherlands", "Belgium", "Sweden", "Czech Republic", "	Portugal", "Greece", "Hungary","Austria", "	Switzerland", "Belarus", "Bulgaria", "Serbia", "Denmark", "Norway", "Finland", "Finland", "Ireland", "Croatia", "	Bosnia and Herzegovina", "Moldova", "Lithuania", "Albania", "Slovenia", "Latvia", "North Macedonia", "Cyprus", "Estonia", "Luxembourg", "Montenegro", "Malta", "Iceland", "Andorra", "Liechtenstein", "Monaco", "San Marino", "Vatican City", "Bosnia and Herzegovina", "Switzerland", "Portugal", "Slovak Republic", "Ukraine", "Holy See") ~ "Europe",
    
    name %in% c("Russia", "Azerbaijan") ~ "Eurasia",
    
    name %in% c("India", "China", "Indonesia", "Pakistan", "Bangladesh", "Japan", "Philippines", "Vietnam", "Iran", "Turkey", "Thailand", "Myanmar", "South Korea", "Iraq", "Afghanistan", "Yemen", "Uzbekistan", "Malaysia", "Saudi Arabia", "Nepal", "North Korea", "Syria", "Sri Lanka", "Kazakhstan", "	Cambodia", "Jordan", "UAE", "Tajikistan", "	Azerbaijan", "Israel", "Laos", "Turkmenistan", "Kyrgyzstan", "Singapore", "Lebanon", "Palestine", "Oman", "Kuwait", "Georgia", "	Mongolia", "Qatar", "Armenia", "Bahrain", "Timor-Leste", "Bhutan", "	Maldives", "Brunei", "Kyrgyz Republic", "Cambodia", "Lao", "Maldives", "Mongolia") ~ "Asia", 
    
    name %in% c("Nigeria", "Ethiopia", "Egypt", "Congo, Dem. Rep.", "Tanzania", "South Africa", "Kenya", "Sudan", "Uganda", "Algeria", "Angola", "Morocco", "Mozambique", "Ghana", "Madagascar", "Cameroon", "Niger", "Mali", "Burkina Faso", "Malawi", "Zambia", "Chad", "Somalia", "Senegal", "Zimbabwe", "Guinea", "Benin", "Rwanda", "Burundi", "South Sudan", "Tunisia", "Togo", "Sierra Leone", "Libya", "Congo, Rep.", "Liberia", "Central African Republic", "Mauritania", "Eritrea", "Namibia", "Gambia", "Gabon", "Botswana", "Lesotho", "Guinea-Bissau", "Equatorial Guinea", "Eswatini", "	Mauritius", "Djibouti", "Comoros", "Cape Verde", "Sao Tome and Principe", "Seychelles", "Cote d'Ivoire", "Mauritius") ~ "Africa"
  ))
```




```{r}
gapminder %>% 
  filter(year== "2022") %>% 
  ggplot(aes(x= gdp, y= cild_death ))+
    geom_point(aes(colour= continent))+
  scale_y_log10()+
  scale_x_log10()+
    geom_smooth(method = lm, rm=T, legend= NULL)+
  labs(title = "GDP vs Child death",
       x= "GDP", 
       y= "Reported Child Deaths",
       color = "Continent")
 
```

```{r}
popgraph1<-gapminder %>% 
  select(year, name, gdp, population) %>% 
  filter(name!= "Monaco",
         year== "2022") %>% 
  arrange(desc(gdp)) %>% 
  slice_head(n=5)
  
```

```{r}
popgraph2<-gapminder %>% 
  select(year, name, gdp, population) %>% 
  filter(name!= "Monaco"&
         name!= "Liechtenstein"&
         name!= "Holy See",
         year== "2022") %>% 
  arrange(desc(gdp)) %>% 
  slice_tail(n=5)
```

```{r}
popgraph<-full_join(popgraph1, popgraph2, by= c("year", "name", "gdp", "population"))
```

```{r}
popgraph %>% 
  ggplot(aes(x=name, y= population))+
  geom_col(aes(fill = name))+
    theme(axis.text.x= element_text(angle = 45, vjust= 0.5))+
  labs(title = "Population of the top 5 and bottom 5 GDP countries",
       x="Country",
       y="Population",
       fill= "country")
```
```{r}
popgraph %>% 
  ggplot(aes(x=name, y= gdp))+
  geom_col(aes(fill = name))+
  theme(axis.text.x= element_text(angle = 45, vjust = 0.5))+
  labs(title = "Countries with the lowest and highest GDP",
       x="Country", 
       y= "GDP",
       fill= "Country")
```
```{r}
gapminder %>% 
  group_by (continent) %>% 
  summarize(mean(life_expectancy, na.rm = T))
```


##Add coordinates or each continent 




Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
