---
title: "Gapminder"
output: html_document
date: "2026-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Libraries
```{r}
library(tidyverse)#includes ggplot2, dplyr, tidyr, readr
library(janitor)#cleaning column names
library(naniar)#missing data visualization
```

```{r}
getwd() #checking the work directory (helps debug file path issues)
```

#Data set and cleaning (Mia)


```{r}
#loading all of the data sets 
gapminder_life_expectancy<- read.csv("data/Life_expectancy.csv") 
```
```{r}
getwd

```

```{r}
co2_emissions<-read_csv("data/co2_pcap_cons.csv")
```

```{r}

Pop<-read_csv("Data/pop.csv")
```

```{r}
Hiv_death<-read_csv("Data/hiv_death.csv")
```

```{r}
Child_death<- read.csv("Data/number_of_child_deaths.csv")

```

```{r}
gdp<- read_csv("Data/gdp.csv")
```
```{r}
lung_cancerF<- read_csv("Data/lung_cancer_f.csv")
```

```{r}
lung_cancerM<- read_csv("Data/lung_cancer_m.csv")
```


#pivoting and fixing year + other problems (mia)

```{r}
#converts wider to long formart (years become rows instead of collumns)
gapminder_life_expectancy<- gapminder_life_expectancy %>% 
  pivot_longer(-c(name,geo),
               names_to = "year",
               values_to = "life_expectancy") %>% 
  mutate(year= str_remove(year, "X"))#removes prefix x from year
```

```{r}
#repeat the pivoting for all datasets
Child_death<-Child_death %>% 
pivot_longer(-c(name,geo),
               names_to = "year",
               values_to = "child_death") %>% 
  mutate(year= str_remove(year, "X")) 
```

```{r}
co2_emissions<- co2_emissions %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
              values_to = "Co2_emissions")
```

```{r}
pop<- Pop %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
              values_to = "population")
```

```{r}
hiv_death<- Hiv_death %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "hiv_deaths")
```

```{r}
gdp<- gdp %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "gdp")
```

```{r}
 lung_cancerF<-lung_cancerF %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "lung_cancerf")
```
  
```{r}
 lung_cancerM<-lung_cancerM %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "lung_cancerm")
```

#Joining data tables

```{r}
#full joins of each loaded dartaser (keeps all data even if missing)
gp<-full_join(Child_death, gapminder_life_expectancy, by= c("name", "geo","year"))
```

```{r}
gp2<-full_join(gp, co2_emissions, by= c("name", "geo","year"))
```

```{r}
gp3<-full_join(gp2,pop, by= c("name", "geo","year"))
```

```{r}
gp4<-full_join(gp3,hiv_death, by= c("name", "geo", "year") )
```

```{r}
#gp5<- full_join(gp4, kcals_per_day, by = c("name","geo","year"))
```

```{r}
gp5<- full_join(gp4, gdp, by = c("name", "year","geo") )
```

```{r}
gp6<- full_join(gp5,lung_cancerF, by = c("name", "year","geo") )
```

```{r}
#the final join to have every dataset together int he tables 
gapminder<- full_join(gp6,lung_cancerM, by = c("name", "year","geo") ) 
```

#FILTERING THE YEARS (mia)
```{r}
gapminder<- gapminder %>% 
  filter(year>= "1915", year<="2022") #filtering the dates to have a more manageable set to work with
```
#seperating each country into a continent (joanna)
```{r}
#assigning continent manually
gapminder <- gapminder %>% 
  mutate(continent = case_when(
    name %in% c("Argentina", "Bolivia", "Brazil", "Chile", "Colombia", "Ecuador", "Guyana", "Paraguay", "Peru", "Suriname", "Uruguay", "Venezuela") ~ "South America",
    
    name %in% c("Australia", "Papua New Guinea", "New Zealand", "Fiji", "Solomon Islands", "Vanuatu", "Samoa", "Kiribati", "Micronesia", "Tonga", "Marshall Islands", "Palau", "Nauru", "Tuvalu", "Micronesia, Fed. Sts.") ~ "Oceania",
    
    name %in% c("USA", "Mexico", "Canada", "Guatemala", "Haiti", "Dominican Republic", "Honduras", "Cuba", "Nicaragua", "El Salvador", "Costa Rica", "Panama", "Jamaica", "Trinidad and Tobago", "Belize", "Bahamas", "Barbados", "St. Lucia", "Grenada", "St. Vincent and the Grenadines", "Antigua and Barbuda", "Dominica", "St. Kitts and Nevis") ~ "North America",
    
    name %in% c("Germany", "UK", "France", "Italy","Spain", "	Ukraine", "Poland", "Romania", "Netherlands", "Belgium", "Sweden", "Czech Republic", "	Portugal", "Greece", "Hungary","Austria", "	Switzerland", "Belarus", "Bulgaria", "Serbia", "Denmark", "Norway", "Finland", "Finland", "Ireland", "Croatia", "	Bosnia and Herzegovina", "Moldova", "Lithuania", "Albania", "Slovenia", "Latvia", "North Macedonia", "Cyprus", "Estonia", "Luxembourg", "Montenegro", "Malta", "Iceland", "Andorra", "Liechtenstein", "Monaco", "San Marino", "Vatican City", "Bosnia and Herzegovina", "Switzerland", "Portugal", "Slovak Republic", "Ukraine", "Holy See") ~ "Europe",
    
    name %in% c("Russia", "Azerbaijan") ~ "Eurasia",
    
    name %in% c("India", "China", "Indonesia", "Pakistan", "Bangladesh", "Japan", "Philippines", "Vietnam", "Iran", "Turkey", "Thailand", "Myanmar", "South Korea", "Iraq", "Afghanistan", "Yemen", "Uzbekistan", "Malaysia", "Saudi Arabia", "Nepal", "North Korea", "Syria", "Sri Lanka", "Kazakhstan", "	Cambodia", "Jordan", "UAE", "Tajikistan", "	Azerbaijan", "Israel", "Laos", "Turkmenistan", "Kyrgyzstan", "Singapore", "Lebanon", "Palestine", "Oman", "Kuwait", "Georgia", "	Mongolia", "Qatar", "Armenia", "Bahrain", "Timor-Leste", "Bhutan", "	Maldives", "Brunei", "Kyrgyz Republic", "Cambodia", "Lao", "Maldives", "Mongolia") ~ "Asia", 
    
    name %in% c("Nigeria", "Ethiopia", "Egypt", "Congo, Dem. Rep.", "Tanzania", "South Africa", "Kenya", "Sudan", "Uganda", "Algeria", "Angola", "Morocco", "Mozambique", "Ghana", "Madagascar", "Cameroon", "Niger", "Mali", "Burkina Faso", "Malawi", "Zambia", "Chad", "Somalia", "Senegal", "Zimbabwe", "Guinea", "Benin", "Rwanda", "Burundi", "South Sudan", "Tunisia", "Togo", "Sierra Leone", "Libya", "Congo, Rep.", "Liberia", "Central African Republic", "Mauritania", "Eritrea", "Namibia", "Gambia", "Gabon", "Botswana", "Lesotho", "Guinea-Bissau", "Equatorial Guinea", "Eswatini", "	Mauritius", "Djibouti", "Comoros", "Cape Verde", "Sao Tome and Principe", "Seychelles", "Cote d'Ivoire", "Mauritius") ~ "Africa"
  ))
```

#add coordinates for mapping (Joanna)
```{r}
#assigning approximate continent center coordinates
gapminder <- gapminder %>% 
  mutate(longitude = case_when(
    continent %in% "Africa" ~ "19.5687",
    continent %in% "Asia" ~ "87.3311",
    continent %in% "Eurasia" ~ "70.5158",
    continent %in% "Europe" ~ "25.3200",
    continent %in% "North America" ~ "-100.1667",
    continent %in% "Oceania" ~ "133.2102",
    continent %in% "South America" ~ "-56.1015"
  ))
```

```{r}
gapminder <- gapminder %>% 
  mutate(latitude = case_when(
    continent %in% "Africa" ~ "6.614",
    continent %in% "Asia" ~ "48.35",
    continent %in% "Eurasia" ~ "49.5786",
    continent %in% "Europe" ~ "54.9067",
    continent %in% "North America" ~ "48.1667",
    continent %in% "Oceania" ~ "-25.9458",
    continent %in% "South America" ~ "-15.5991"
  ))
```

```{r}
gapminder <- gapminder %>% 
  mutate(longitude = as.numeric(longitude),
         latitude = as.numeric(latitude))
```


#final data set

```{r}
gapminder
```

```{r}
#dataset written as csv in order to open easier in shiny apps
write.csv(gapminder, file="gapminder.csv", row.names=FALSE)
```

```{r}
gapminder<-read_csv("gapminder.csv")
#gapminder  (checking the dataset looks good)
```

#visualiding missing values (mia)
```{r}
gg_miss_var(gapminder)
```
#summary of missing data (mia)
```{r}
gapminder %>% 
  miss_var_summary(order=T)
```


#Analysis examples

?is there correlation between the top and bottom co2 emissions and lung cancer, does it look as though there is enough data to see?

#top 3 highest co2 emissions per capita (mia)

```{r}
gapminder %>% 
  select(name, Co2_emissions, year) %>% 
  pivot_wider(names_from = "name",
              values_from = "Co2_emissions")
```


```{r}
gapminder %>% 
  filter(year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(avgco2=mean(Co2_emissions)) %>% 
  arrange(desc(avgco2)) %>% 
  slice_head(n=3)
  
```

#bottom 3 lowest co2 emissions (mia)
```{r}
gapminder %>% 
  filter(name != c("Holy See", "Monaco", "San Marino"), year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(avgco2=mean(Co2_emissions, na.rm=T)) %>% 
  arrange(desc(avgco2)) %>% 
slice_tail(n=6)

```

#6 coutnries and the lung cancer deaths (mia)

```{r}
gapminder %>% 
   filter(name ==c("Chad", "Congo, Dem. Rep.", "Burundi"), year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(mlungavg= mean(lung_cancerm), flungavg=mean(lung_cancerf)) 
```

#highest year of co2 (mia)
```{r}
gapminder %>%
  select(name, year, Co2_emissions) %>% 
  filter(year >= "1980", year<="2022") %>% 
pivot_wider(names_from = "year",
            values_from = Co2_emissions) 
  
```
#add total lung cancer (mia)
```{r}
gapminder<- gapminder %>% 
  mutate(lungcancertot=lung_cancerf+lung_cancerm)
```

#co2 emissions between the 3 largest states in 2000 (mia)

```{r}
gapminder %>%
  select(name, year, Co2_emissions, lung_cancerf, lung_cancerm) %>% 
  filter(name== "Qatar"|
         name == "Singapore"|
         name == "USA"|
           name== "Chad"|
           name== "Burundi"|
           name== "Congo, Dem. Rep."|
         name=="UAE",
         year== "2000") %>% 
   ggplot(aes(x=name, y=Co2_emissions, fill = name ))+
  geom_col()+
  scale_y_log10()+
  labs(title = "Co2 Emissions in the year 2000",
       x="Country",
       y= "Co2 Emissions(per capita)")
    
```

#lung cancer by sex (mia)

```{r}
gapminder %>%
  select(name, year, Co2_emissions, lungcancertot, lung_cancerf, lung_cancerm) %>% 
  filter(name== "Qatar"|
         name == "Singapore"|
         name == "USA"|
           name== "Chad"|
           name== "Burundi"|
         name=="UAE",
         year =="2000") %>%  #qatar, singapore, usa, chad, burundi
   pivot_longer(cols = c(lung_cancerf, lung_cancerm),
               names_to = "sex",
               values_to = "deaths")  %>% 
  ggplot(aes(x=name, y= deaths, fill = sex ))+
  geom_col(position= "dodge")+
  scale_y_log10()+
  labs(title = "Lung cancer deaths by sex in the year 2000",
       x="Country",
       y= "Deaths",
       fill= "Sex")
  

```

trying to see correlation between gdp and the amoutn of child deaths, ato later see if gdp and population have a similar or different relationship. 

#gdp vs child deaths (mia)


```{r}
gapminder %>% 
  distinct(name) %>% 
  arrange(name)
```

```{r}
gapminder %>% 
  filter(year== "2022") %>% 
  ggplot(aes(x= gdp, y= child_death ))+
    geom_point(aes(colour= continent))+
  scale_y_log10()+
  scale_x_log10()+
    geom_smooth(method = lm, rm=T, legend= NULL)+
  labs(title = "GDP vs Child death",
       x= "GDP", 
       y= "Reported Child Deaths",
       color = "Continent")
 
```
Here is the work we did to see if a population siz emight have an increase or a larger amount when gdp is higher, becuase child death is lower

#g=creating smaller dataset (mia)

```{r}
#creating a new data set in order to quickly plot the graphs for population vs gdp of the top 10 countries
popgraph1<-gapminder %>% 
  select(year, name, gdp, population) %>% 
  filter(name!= "Monaco",
         year== "2022") %>% 
  arrange(desc(gdp)) %>% 
  slice_head(n=5)
  
```

```{r}
#second half of the smaller dataset
popgraph2<-gapminder %>% 
  select(year, name, gdp, population) %>% 
  filter(name!= "Monaco"&
         name!= "Liechtenstein"&
         name!= "Holy See",
         year== "2022") %>% 
  arrange(desc(gdp)) %>% 
  slice_tail(n=5)
```

```{r}
#joining the two datasets in order to have one graph fro the poulations
popgraph<-full_join(popgraph1, popgraph2, by= c("year", "name", "gdp", "population"))
```
#popualtion graph (mia)

```{r}
popgraph %>% 
  ggplot(aes(x=name, y= population))+
  geom_col(aes(fill = name))+
    theme(axis.text.x= element_text(angle = 45, vjust= 0.5))+
  labs(title = "Population of the top 5 and bottom 5 GDP countries",
       x="Country",
       y="Population",
       fill= "country")
```

```{r}
popgraph %>% 
  ggplot(aes(x=name, y= gdp))+
  geom_col(aes(fill = name))+
  theme(axis.text.x= element_text(angle = 45, vjust = 0.5))+
  labs(title = "Countries with the lowest and highest GDP",
       x="Country", 
       y= "GDP",
       fill= "Country")
```
```{r}
gapminder %>% 
  group_by (continent) %>% 
  summarize(mean(life_expectancy, na.rm = T))
```







Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
