---
title: "Gapminder"
output: html_document
date: "2026-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Libraries
```{r}
library(tidyverse)
library(tibble)
library(janitor)
library(naniar)
library(dplyr)
library(shiny)
library(shinydashboard)
library(shinythemes)
```

#Data set and cleaning

```{r}
gapminder_life_expectancy<- read.csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/Life_expectancy.csv") 
```

```{r}
co2_emissions<-read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/co2_pcap_cons.csv")
```

```{r}
Pop<-read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/pop.csv")
```

```{r}
Hiv_death<-read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/hiv_death.csv")
```
```{r}
Child_death<- read.csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/number_of_child_deaths.csv")

```

```{r}
kcals_per_day<- read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/kilocalories_per_person_and_day.csv")
```

```{r}
gdp<- read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/gdp.csv")
```
```{r}
lung_cancerF<- read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/lung_cancer_f.csv")
```

```{r}
lung_cancerM<- read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/Data/lung_cancer_m.csv")
```


#pivoting and fixing year + other problems

```{r}
gapminder_life_expectancy<- gapminder_life_expectancy %>% 
  pivot_longer(-c(name,geo),
               names_to = "year",
               values_to = "life_expectancy") %>% 
  mutate(year= str_remove(year, "X"))
```

```{r}
Child_death<-Child_death %>% 
pivot_longer(-c(name,geo),
               names_to = "year",
               values_to = "cild_death") %>% 
  mutate(year= str_remove(year, "X")) 
```

```{r}
co2_emissions<- co2_emissions %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
              values_to = "Co2_emissions")
```

```{r}
pop<- Pop %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
              values_to = "population")
```

```{r}
hiv_death<- Hiv_death %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "hiv_deaths")
```

```{r}
kcals_per_day<- kcals_per_day %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "kcal")
```

```{r}
gdp<- gdp %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "gdp")
```

```{r}
 lung_cancerF<-lung_cancerF %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "lung_cancerf")
```
  
```{r}
 lung_cancerM<-lung_cancerM %>% 
  pivot_longer(-c(geo,name),
               names_to = "year",
               values_to = "lung_cancerm")
```

#joining data tables

```{r}
gp<-full_join(Child_death, gapminder_life_expectancy, by= c("name", "geo","year"))
```

```{r}
gp2<-full_join(gp, co2_emissions, by= c("name", "geo","year"))
```

```{r}
gp3<-full_join(gp2,pop, by= c("name", "geo","year"))
```

```{r}
gp4<-full_join(gp3,hiv_death, by= c("name", "geo", "year") )
```

```{r}
gp5<- full_join(gp4, kcals_per_day, by = c("name","geo","year"))
```

```{r}
gp6<- full_join(gp5, gdp, by = c("name", "year","geo") )
```

```{r}
gp7<- full_join(gp6,lung_cancerF, by = c("name", "year","geo") )
```

```{r}
gapminder<- full_join(gp7,lung_cancerM, by = c("name", "year","geo") )
```


```{r}
gapminder<- gapminder %>% 
  filter(year>= "1915", year<="2022")
```


#final data set

```{r}
gapminder
```

```{r}
write.csv(gapminder, file="gapminder.csv", row.names=FALSE)
```

```{r}
gapminder<-read_csv("C:/Users/Miala/OneDrive/Desktop/BIS15L_W2026_Group9/Gapminder/gapminder.csv")
gapminder
```

#understanding the data
```{r}
gg_miss_var(gapminder)
```

```{r}
gapminder %>% 
  miss_var_summary(order=T)
```


```{r}
gapminder %>% 
  select(name, Co2_emissions, year) %>% 
  pivot_wider(names_from = "name",
              values_from = "Co2_emissions")
```


#graphs
top 3 highest co2 emissions per capita
```{r}
gapminder %>% 
  filter(year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(avgco2=mean(Co2_emissions)) %>% 
  arrange(desc(avgco2)) %>% 
  slice_head(n=3)
  
```
bottom 3 lowest co2 emissions
```{r}
gapminder %>% 
  filter(name != c("Holy See", "Monaco", "San Marino"), year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(avgco2=mean(Co2_emissions, rm.na=T)) %>% 
  arrange(desc(avgco2)) %>% 
slice_tail(n=6)

```

6 coutnries and the lung cancer deaths

```{r}
gapminder %>% 
   filter(name ==c("Chad", "Congo, Dem. Rep.", "Burundi"), year>= "1980", year<="2022") %>% 
  group_by(name) %>% 
  summarize(mlungavg= mean(lung_cancerm), flungavg=mean(lung_cancerf)) %>% 
  ggplot(aes(x=name, y= ))
 
```

highest year of co2
```{r}
gapminder %>%
  select(name, year, Co2_emissions) %>% 
  filter(year >= "1980", year<="2022") %>% 
pivot_wider(names_from = "year",
            values_from = Co2_emissions) 
  
```

```{r}
gapminder<- gapminder %>% 
  mutate(lungcancertot=lung_cancerf+lung_cancerm)
```

co2 emissions between the 3 largest states in 2000

```{r}
gapminder %>%
  select(name, year, Co2_emissions, lung_cancerf, lung_cancerm) %>% 
  filter(name== "Qatar"|
         name == "Singapore"|
         name == "USA"|
           name== "Chad"|
           name== "Burundi"|
           name== "Congo, Dem. Rep."|
         name=="UAE",
         year== "2000") %>% 
   ggplot(aes(x=name, y=Co2_emissions, fill = name ))+
  geom_col()+
  scale_y_log10()+
  labs(title = "Co2 Emissions in the year 2000",
       x="Country",
       y= "Co2 Emissions(per capita)")
    
```



```{r}
gapminder %>%
  select(name, year, Co2_emissions, lungcancertot, lung_cancerf, lung_cancerm) %>% 
  filter(name== "Qatar"|
         name == "Singapore"|
         name == "USA"|
           name== "Chad"|
           name== "Burundi"|
         name=="UAE",
         year =="2000") %>%  #qatar, singapore, usa, chad, burundi
   pivot_longer(cols = c(lung_cancerf, lung_cancerm),
               names_to = "sex",
               values_to = "deaths")  %>% 
  ggplot(aes(x=name, y= deaths, fill = sex ))+
  geom_col(position= "dodge")+
  scale_y_log10()+
  labs(title = "Lung cancer deaths by sex in the year 2000",
       x="Country",
       y= "Deaths",
       fill= "Sex")
  

```
```{r}
gapminder
```

```{r}

```


```{r}
gapminder %>% 
  filter(year== "2022") %>% 
  ggplot(aes(x= gdp, y= cild_death ))+
    geom_point(aes(colour= name))+
  scale_y_log10()+
  scale_x_log10()+
    geom_smooth(method = lm, rm=T, legend= NULL)+
  labs(title = "GDP vs Child death",
       x= "GDP", Y= "Reported Child Deaths")+
  theme(legend.position= "none")
 
```

```{r}
popgraph1<-gapminder %>% 
  select(year, name, gdp, population) %>% 
  filter(name!= "Monaco",
         year== "2022") %>% 
  arrange(desc(gdp)) %>% 
  slice_head(n=5)
  
```

```{r}
popgraph2<-gapminder %>% 
  select(year, name, gdp, population) %>% 
  filter(name!= "Monaco"&
         name!= "Liechtenstein"&
         name!= "Holy See",
         year== "2022") %>% 
  arrange(desc(gdp)) %>% 
  slice_tail(n=5)
```

```{r}
popgraph<-full_join(popgraph1, popgraph2, by= c("year", "name", "gdp", "population"))
```

```{r}
popgraph %>% 
  ggplot(aes(x=name, y= population))+
  geom_col(aes(fill = name))+
    theme(axis.text.x= element_text(angle = 45))+
  labs(title = "Population of the top 5 and bottom 5 GDP countries",
       x="Country",
       Y="Population",
       fill= "Country")
```
```{r}
popgraph %>% 
  ggplot(aes(x=name, y= gdp))+
  geom_col(aes(fill = name))+
  theme(axis.text.x= element_text(angle = 45))+
  labs(title = "Countries with the lowest and highest GDP",
       x="Country", 
       y= "GDP",
       fill= "Country")
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
